### 2021-07-04 abs.tex
rolling on the conclusion

This thesis presents four projects completed in the course of my PhD which address these problems.
First, I demonstrate how to calculate the sediment flux from the dynamics of individual grains and describe the trajectories of grains alternating through rests and motions having fluctuating velocities, linking the sediment flux to the particle dynamics and describing these with more detail.
Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting the residence times of particles buried within the bed.
Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
Finally I describe particles propelled downstream through episodic collisions with the bed surface with a granular gas-like approach, fitting all experimental data on downstream sediment velocities.
I conclude by summarizing these developments, discussing their implications for the statistical mechanics sediment transport framework, and suggesting how we can use this progress to better understand landscapes.

>>>>

This thesis presents four projects completed in the course of my PhD which address these problems.
First, I demonstrate how to calculate the sediment flux from the dynamics of individual grains and describe the trajectories of grains alternating through rests and motions having fluctuating velocities, linking the sediment flux to the particle dynamics and describing these with more detail.
Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting the residence times of particles buried within the bed.
Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
Finally I describe particles propelled downstream through episodic collisions with the bed surface submissionusing a granular gas-like theory, fitting all experimental data on downstream sediment velocities.
I conclude by summarizing these developments, discussing their implications for the statistical mechanics sediment transport framework, and suggesting how we can use this progress to better understand landscapes.

### 2021-07-04 app_ch3.tex
made some progress on the nightmare citations.
rolling on the conclusion

\chapter{Ch. 3 calculations}

@@

The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1992} and \citet{Gillespie2007}.
The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
In the following sections, we begin by demonstrating the time intervals $\tau$ between subsequent transitions (i.e., the intervals between transition times) are exponentially distributed within the model we develop in the main text. Then we describe the SSA as a consequence of this property.

>>>>

\chapter{Calculations involved in the bed elevation model}

@@

The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1991} and \citet{Gillespie2007}.
The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
In the following sections, we begin by demonstrating the time intervals $\tau$ between subsequent transitions (i.e., the intervals between transition times) are exponentially distributed within the model we develop in the main text. Then we describe the SSA as a consequence of this property.

### 2021-07-04 ch1intro.tex
rolling on the conclusion

\subsection{Motivation: tracers and basic understanding}
An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013}. This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.

\subsection{Einstein 1937}
Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}.
Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to probabilistic methods to characterize their transport.
Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
As his interest was on the global range of particle motion, and the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
With this configuration, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed. In the process, Einstein formulated an example of what is now called the continuous time random walk, but this was not was not formalized until much later \citep{Montroll1965}.

@@

For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
\be  \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) \ee
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase. The variance picks up a two range scaling:
Computing $\sigma_x^2$ provides
\be \sigma_x^2(t) = ,\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, since fluctuations of the velocity between entrainment and deposition have been neglected.

@@

\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance. The timescale $t_r$ is a relaxation time over which velocity fluctuations decay. This time, the Fokker-Planck equation is
\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
and the steady state solution becomes
\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
which is the Gaussian velocity distribution from the other subset of the experiments.

These two models of bedload velocity distributions are summarized in figure \ref{fig:fanAncey}.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag. Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d). The latter has relatively narrow and symmetric fluctuations, with the mode well above $u=0$.}
	\label{fig:fanAncey}
\end{figure}

\subsection{The definition of the flux}

Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
Today, there are two main competing (or complementary) approaches to define the bedload flux. 
The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\mathcal{S}$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\mathcal{\textbf{S}}. \ee
This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.
The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $\mathcal{V}$:
\be q = \frac{1}{L} \sum_{i\in\mathcal{V}} u_i. \label{eq:controlflux}\ee
Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.

@@

Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
Einstein formulated his original model of individual grains in transport \citep{Einstein1937} in terms of particle entrainment and deposition using the conceptual picture of grains moving downstream through a sequence of instantaneous steps.
Later, \citet{Einstein1942,Einstein1950} calculated the bulk sediment flux with these same probabilistic ideas, providing an alternative to the Bagnold scaling approach.
The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
 \begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
left to right through an array of adjacent control volumes. The bedload flux is the rate of bedload particles crossing
the surface $\mathcal{S}$ per unit width and time, collected from all upstream control volumes.}
	\label{fig:einsteinFluxConcept}
\end{figure}

He partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
Each control volume has downstream length $\ell$ which is also the average particle step length.
Denoting by $P_n$ the probability that an individual grain undergoes at least $n$ jumps of length $\ell$ in a time interval $T$, meaning it travels at least a distance $n \ell$, and considering that there is a density $\rho$ of particles at rest on the bed, it follows that on average $\rho \ell P_n$ particles will displace a distance $n \ell$ or more from within each control volume during the time interval $T$. 
As a result, since grains crossing $\mathcal{S}$ in a time $T$ could have come from any upstream location, the number of grains crossing $\mathcal{S}$ in $T$ is a sum over all control volumes: $\sum_{n=1}^\infty \rho \ell P_n$.
Dividing by the time $T$ to get the average rate of grains crossing $\mathcal{S}$ provides the mean flux:
\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty P_n. \label{eq:einflux} \ee
The final quantity to evaluate is $P_n$, the probability a particle entrains \textit{at least} $n$ times in a time $T$.

@@

\citet{Nakagawa1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation based upon Einstein's concepts of entrainment and deposition rates.
According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition:
\be \partial_x q(x,t) = E - D. \ee 
Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described in an Einstein-like framework: 
\be (1-\phi) px z(x,t) = D - E.\ee
In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the result of entrainment at all upstream locations, giving
\be (1-\phi) px z(x,t) = \int_0^\infty dx' int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.

@@

Chapter \ref{ch:burial} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movements characteristics of grains.

@@

Chapter \ref{ch:downdiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.

>>>>

\subsection{Motivation: tracers and basic understanding}
An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}. This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.

\subsection{Einstein's random walk model}
Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}.
Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to probabilistic methods to characterize their transport.
Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
As his interest was on the global range of particle motion, and the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
With this configuration, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed. In the process, Einstein formulated an example of what is now called the continuous time random walk, but this was not was not formalized until much later \citep{Montroll1965}.

@@

For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
\be  P(x,t) =  \ee
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase. The variance picks up a two range scaling:
Computing $\sigma_x^2$ provides
\be \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) ,\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, since fluctuations of the velocity between entrainment and deposition have been neglected.

@@

\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance. The timescale $t_r$ is a relaxation time over which velocity fluctuations decay. This time, the Fokker-Planck equation is
\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
and the steady state solution becomes
\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
which is the Gaussian velocity distribution from the other subset of the experiments.
These two models of bedload velocity distributions are summarized in figure \ref{fig:fanAncey}.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag. Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d). The latter has relatively narrow and symmetric fluctuations, with the mode well above $u=0$.}
	\label{fig:fanAncey}
\end{figure}

\subsection{The definition of the flux}
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.pdf}
	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$. Here, particles are shown moving from time $t-dt$ (faint white particles) to time $t$ (grey particles). Each particle's velocity component $u_i$ in the downstream direction is indicated. At the instant $t$, the surface definition of the flux \ref{eq:surfflux} gives $q = \frac{1}{L}(u_7+u_8).$, while the volume definition \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_6)$,
	showing the non-equivalence of these two definitions.}
	\label{fig:fluxdefs}
\end{figure}
Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
Today, there are two main competing (or complementary) approaches to define the bedload flux. 
The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\mathcal{S}$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\mathcal{\textbf{S}}. \label{eq:surfflux} \ee
This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.
The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $V$:
\be \Phi = \frac{1}{L} \sum_{i\in V} u_i. \label{eq:controlflux}\ee
Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.

@@

Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
Einstein formulated his original model of individual grains in transport \citep{Einstein1937} in terms of particle entrainment and deposition using the conceptual picture of grains moving downstream through a sequence of instantaneous steps.
Later, \citet{Einstein1942,Einstein1950} calculated the bulk sediment flux with these same probabilistic ideas, providing an alternative to the Bagnold scaling approach.
The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
 \begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
left to right through an array of adjacent control volumes. The bedload flux is the rate of bedload particles crossing
the surface $S$ per unit width and time, collected from all upstream control volumes.}
	\label{fig:einsteinFluxConcept}
\end{figure}

He partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
Each control volume has downstream length $\ell$ which is also the average particle step length.
Denoting by $P_n$ the probability that an individual grain undergoes at least $n$ jumps of length $\ell$ in a time interval $T$, meaning it travels at least a distance $n \ell$, and considering that there is a density $\rho$ of particles at rest on the bed, it follows that on average $\rho \ell P_n$ particles will displace a distance $n \ell$ or more from within each control volume during the time interval $T$. 
As a result, since grains crossing $S$ in a time $T$ could have come from any upstream location, the number of grains crossing $S$ in $T$ is a sum over all control volumes: $\sum_{n=1}^\infty \rho \ell P_n$.
Dividing by the time $T$ to get the average rate of grains crossing $S$ provides the mean flux:
\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty P_n. \label{eq:einflux} \ee
The final quantity to evaluate is $P_n$, the probability a particle entrains \textit{at least} $n$ times in a time $T$.

@@

\citet{Nakagawa1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation based upon Einstein's concepts of entrainment and deposition rates.
According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition:
\be \partial_x q(x,t) = E - D. \ee 
Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described in Einstein's variables: 
\be (1-\phi) \px z(x,t) = D - E.\ee
In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the result of entrainment at all upstream locations, giving
\be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.

@@

Chapter \ref{ch:ch3} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movements characteristics of grains.

@@

Chapter \ref{ch:downDiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.

### 2021-07-04 ch5langevin.tex
rolling on the conclusion

\chapter{Collisional Langevin model of the bedload sediment velocity distributions}
\label{ch:langevin}
\section{Introduction}

Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Due to these fluctuations, measured transport rates often show slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude \citep{Ancey2020}.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{}.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations \citep{Ancey2008}, quantifying the dependence of measurements on the observation scale \citep{Turowski2010}, and linking bulk transport characteristics to the ``microstructural" dynamics of individual grains \citep{Ancey2014}.
Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume \citep{Ancey2008,Furbish2012}.
These approaches rely on the instantaneous velocity distribution of sediment particles \citep{Lajeunesse2010}, but our understanding of these distributions unfortunately remains limited.
We have as of yet no consensus on the shape of the bedload velocity distribution (is it Exponential, Gaussian, Gamma, or some other shape?), and although models have described some extreme endmember distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.

@@

With the above forces, we express the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics as
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier stocahstic bed load models with an episodic term which provides a more realistic representation of particle-bed collisions during saltation.
It represents a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulent generates "diffusion".
The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice since the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional.
Given this integration rule, the remaining issues are to obtain the integro-differential equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

>>>>

\chapter{Collisional Langevin model of bedload sediment velocity distributions}
\label{ch:langevin}
\section{Introduction}

Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Due to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations \citep{Ancey2008} and quantifying the dependence of measurements on the observation scale \citep{Turowski2010}. 
Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume \citep{Ancey2008,Furbish2012}.
This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although our understanding of these distributions unfortunately remains limited.
We have as of yet no consensus on the shape of the bedload velocity distribution, and although models have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.

@@

With the above forces, we express the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics as
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier stocahstic bed load models with an episodic term which provides a more realistic representation of particle-bed collisions during saltation.
It represents a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulent generates "diffusion".
The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2016,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice since the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional.
Given this integration rule, the remaining issues are to obtain the integro-differential equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

### 2021-07-04 ch6conc.tex
rolling on the conclusion

\chapter{Summary and future work}
\label{ch:conc}
\section{The overall strategy}
\subsection{Langevin and master equations}
\subsection{Idealized noises and their combinations}

@@

\subsection{Probability distribution of the sediment flux from micromechanics of particle trajectories}

@@

\subsection{Predicting how sediment burial affects the downstream spreading of sediment tracer particles}

\section{Models and the real world}

\section{Conclusion and next steps}

>>>>

\chapter{Summary and future work}
\label{ch:conc}
\section{Overall methodology of the thesis}

The research in this thesis generalizes and extends earlier work to describe sediment transport as a stochastic process. In the research, I have characterized the movements of grains along a wider range of timescales than before, framed existing research in terms of Langevin and Master equations, and produced more detailed descriptions of the bedload transport phenomenon, addressing several long-recognized shortcomings of earlier stochastic models. In particular I have filled in the connection between sediment fluxes and individual praticle motions, included particle velocity fluctuations in a descriptions of individual particle motions through rest and motion, evaluated the role of bed elevation changes on sediment transport rate fluctuations, and included the process of sediment burial in descriptions of downstream particle movement.

\subsection{Langevin and master equations}

The over-arching strategy in all of these extentions has been to identify control and response variables, represent variable controls by idealized noises (entrainment and deposition events, particle burial events, turbulent forces, or particle-bed collisions), and then to write dynamical equations relating the respones variables to these stochastic control variables.
This approach phrases sedimentary dynamics in terms of Langevin-like equations, stochastic analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for a the response variable of interest in the problem and the force $F$ is a stochastic idealization of the control variables.

Once such a stochastic dynamical equation was written for a given problem, its solutions for all realizations of the noise which provide a given outcome were counted () to form a probability distribution of the response variable. These distributions are generally given as solutions to integro-differential equations (), and the distributions were determined as solutions to these equations with transform calculus ().

In every case, the result is a probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (), particle velocity (), bed elevation (), and sediment transport rate ().

\subsection{Idealized noises and their combinations}

A major challenge in this research is that only a handful noises in statistical physics are comprehensively understood \citep{}, so the available options in stochastic modelling are constrained.
The noises used in this thesis include white Gaussian, Poisson, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively. 
Turbulence was described using Gaussian noise (); instantaneous steps (), particle arrivals to a control surface (), and particle-bed collisions () were described with Poisson noise; and alternation between motion and rest was represented with dichotomous noise ().
Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise () as required.
This inclusion of multiple noise sources has not yet to my knowledge been pursued in earlier stochastic models of sediment transport.

River science offers no guarantee that the idealized white noises we understand best are sufficient to describe its phenomena.
White noise spectra are an idealization which is probably never represented in nature \citep{Gardiner1983}.
In contrast, colored spectra in which fluctuations distribute power unevenly across frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of turbulence \citep{Kolmogorov1941,Nikora2000} and avalanches \citep{Bak1987,Jensen1998}.
In river science, many phenomena exhibit colored spectra, like the size distribution of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment fluxes \citep{Dhont2018,Chartrand2021}.
Dynamical equations driven by colored noise are mathematically extremely challenging to solve and require case-by-case analysis \citep{Hanggi1978,Luczka2005,Hanggi2007}.
This means even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.

@@

\subsection{Probability distribution of the sediment flux from micromechanics of particle transport}

@@

\subsection{Predicting how sediment burial affects the downstream transport of sediment particles}

\section{Limitations and future research directions}

\subsection{Morphology-transport interplay}

\subsection{Grain size distributions}

\subsection{Transport stage}

\subsection{Continuum descriptions of sedimentary beds}

\section{Conclusion}

This thesis accepted variability in grain scale bedload transport to form some new descriptions of sediment transport processes in river channels.

The founders of process-based geomorphology always acknowledged variability in geomorphology \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes despite it, such as competent conditions to summarize climatic fluctuations \citep{Wolman1959,Wolman1978}, global characteristics of rivers and basins like channel orders and network topologies to avoid variable local characteristics, and averaged flow conditions to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}.

The stochastic models of sediment transport developed in this thesis demonstrate a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through equations governing landscape evolution, and revisit the old question of how variability shapes Earth's surface.
